---
title: "LAB 2:: HIDDEN MARKOV MODELS"
date: "`r Sys.Date()`"
author: "Johannes Hedström"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A96 Advanced Machine Learning \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\
  Department of Computer and Information Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE, fig.width = 5, fig.height = 3, fig.align = 'center')

```





```{r}

library(HMM)
library(knitr)

```


# Description

The purpose of the lab is to put in practice some of the concepts covered in the lectures.
To do so, you are asked to model the behavior of a robot that walks around a ring. The ring is
divided into 10 sectors. At any given time point, the robot is in one of the sectors and decides
with equal probability to stay in that sector or move to the next sector. You do not have direct
observation of the robot. However, the robot is equipped with a tracking device that you can
access. The device is not very accurate though: If the robot is in the sector i, then the device
will report that the robot is in the sectors [i − 2, i + 2] with equal probability.




# Question 1

Build a hidden Markov model (HMM) for the scenario described above. Note that the
documentation for the initHMM function says that the emissionProbs matrix should
be of size [number of states]x[number of states]. This is wrong. It should be of size
[number of states]x[number of symbols]. The package works correctly, though. It is
just the documentation that is wrong.


```{r}
# States 
states <- as.character(1:10)
# Movements for the robot
sym <- c('Stay','Move up', 'Move down')
# starting probabilities, equal for all states
startprobs <- rep(1/10,10)

# Transitions probabilities 1/3 probability to stay or change

t_prob <- matrix(0,nrow=10, ncol=10) # empty matrix

for (i  in 1:10) { # filling the matrix with probabilities, each row sums to 1
  j <- (i-3):(i+3)
  if(11 %in% j){
    
    j_1 <-j[1:which(j==10)]   
    j <- c(1:(length(j) - length(j_1)),j_1)
    }
  if(0 %in% j){
    j <- c(1,2,10)
  }
  t_prob[i,j] <- 1/3
}


# emissionprobs

e_prob <- matrix()

hmm <-initHMM(states,sym, startprobs,transProbs=t_prob)

hmm
```


# Question 2

Simulate the HMM for 100 time steps.

# Question 3

Discard the hidden states from the sample obtained above. Use the remaining observations to compute the filtered and smoothed probability distributions for each of the
100 time points. Compute also the most probable path.

# Question 4

Compute the accuracy of the filtered and smoothed probability distributions, and of the
most probable path. That is, compute the percentage of the true hidden states that are
guessed by each method.

Hint: Note that the function forward in the HMM package returns probabilities in log
scale. You may need to use the functions exp and prop.table in order to obtain a
normalized probability distribution. You may also want to use the functions apply and
which.max to find out the most probable states. Finally, recall that you can compare
two vectors A and B element wise as A==B, and that the function table will count the
number of times that the different elements in a vector occur in the vector

# Question 5

Repeat the previous exercise with different simulated samples. In general, the smoothed
distributions should be more accurate than the filtered distributions. Why ? In general,
the smoothed distributions should be more accurate than the most probable paths, too.
Why ?

# Question 6

Is it always true that the later in time (i.e., the more observations you have received)
the better you know where the robot is ?
Hint: You may want to compute the entropy of the filtered distributions with the function
entropy.empirical of the package entropy.

# Question 7

Consider any of the samples above of length 100. Compute the probabilities of the
hidden states for the time step 101.


